---
title: "Project Final"
author: "Zain Shaikh"
date: "2024-04-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Predicting Medical Insurance Cost

## Loading data

```{r}
library(ggplot2)
library(ROSE)

med <- read.csv("medical_insurance.csv")

head(med)
str(med)




```

## Data Cleaning

```{r}

nrow(med) # total rows

sum(duplicated(med)) # duplicated rows

med <- med[!duplicated(med), ] #without duplicates

med <- na.omit(med)

tab <- table(med$smoker) #smoker counts

barplot(table(med$smoker)) #Smoker


barplot(table(med$children)) #Children

barplot(table(med$region)) #region

barplot(table(med$sex)) #Sex


# library(ggplot2)
# ggplot(med, aes(x = smoker, charges)) +
#   geom_bar()



# Factorize

med <- within(med, {sex = factor(sex) ; smoker = factor(smoker) ; region = factor(region) ; children = factor(children) })
str(med)





```

## Exploratory Data Analysis

```{r}

pairs(charges ~ . , data = med) #scatterplot matrix


# hist of bmi
ggplot(med, aes(bmi)) +
  geom_histogram(bins = 35,colour="black",fill="plum") + 
  labs(title = "histogram of BMI",x="BMI", y = "Count")


#hist of charges
ggplot(med, aes(charges)) +
  geom_histogram(bins = 25,colour="black",fill="plum") + 
  labs(title = "histogram of Charges",x="Charges", y = "Count")

#(Suggests transformation)


#hist of age
ggplot(med, aes(age)) +
  geom_histogram(bins = 20,colour="black",fill="plum") + 
  labs(title = "histogram of Age",x="Age", y = "Count")







# Load the necessary libraries

```

## Models

```{r}
library(faraway) # to use vif() function
library(car)
library(MASS)

lm_full <- lm(charges ~ . , data = med) #sex Not sigf. Transformation ?
summary(lm_full)

plot(lm_full)




lm_2 <- lm(log(charges) ~ . , data = med) #further improvement needed
summary(lm_2)
plot(lm_2)


round(vif(lm_2), 2) # no multicollinearity




plot(med$age, rstandard(lm_2))
plot(med$bmi, rstandard(lm_2)) # No motivation for polynomial


#checking for interactions for improvement


ggplot(med, aes(bmi, charges, colour = smoker)) +  #No interaction
geom_point(alpha=0.7) +
geom_smooth(method="lm", se=FALSE)

ggplot(med, aes(age + bmi, charges, colour = smoker)) +  #Yes interaction. 3 way
geom_point(alpha=0.7) +
geom_smooth(method="lm", se=FALSE)


ggplot(med, aes(bmi, charges, colour = smoker)) +  #Yes Interaction
geom_point(alpha=0.7) +
geom_smooth(method="lm", se=FALSE)

ggplot(med, aes(bmi, charges, colour = region)) + #Somewhat interaction
geom_point(alpha=0.7) +
geom_smooth(method="lm", se=FALSE)

ggplot(med, aes(bmi, charges, colour = region)) + #somewhat interaction
geom_point(alpha=0.7) +
geom_smooth(method="lm", se=FALSE)


ggplot(med, aes(bmi, charges, colour = children)) + #somewhat interaction. Check
geom_point(alpha=0.7) +
geom_smooth(method="lm", se=FALSE)


ggplot(med, aes(age, charges, colour = children)) + #somewhat interaction
geom_point(alpha=0.7) +
geom_smooth(method="lm", se=FALSE)


ggplot(med, aes(bmi, charges, colour = sex)) + #Yes interaction
geom_point(alpha=0.7) +
geom_smooth(method="lm", se=FALSE)


ggplot(med, aes(age, charges, colour = sex)) + #No Interaction
geom_point(alpha=0.7) +
geom_smooth(method="lm", se=FALSE)



lm_3 <- lm(log(charges) ~ age + bmi + children + sex + region + bmi:smoker + bmi:region + bmi:children + age:children + bmi:sex, data = med)
summary(lm_3)


lm_4 <- lm(log(charges) ~ age + bmi + children + sex + region + smoker + bmi:smoker + bmi:region + age:children + age:smoker:bmi, data = med)
summary(lm_4)


plot(lm_4)


lm_4_step <- step(lm_4) # Same model as ours
summary(lm_4_step)

AIC(lm_4_step)



AIC(lm_full) #Worst model
AIC(lm_2)
AIC(lm_4)  # Best model with lowest



summary(powerTransform(lm_4))
boxcox(lm_4, lambda = seq(1, 3, 0.05))




lm_5 <- lm(log(charges)^2 ~ age + bmi + children + sex + region + bmi:smoker + bmi:region + age:children + age:smoker:bmi, data = med)
summary(lm_5)

plot(lm_5)

AIC(lm_5) # checked box cox but AIC went up


plot(fitted(lm_4), med$charges) # observed vs fitted
plot(fitted(lm_full), med$charges)

```

Equation:

$log(Charges) = 7.0034 + 0.0400Age + 0.0017BMI + 0.2571Children1 + 0.8016Children2 + 0.6053Children3 + 1.1510Children4 + 0.9085Children5 - 0.0921Sex - 0.0082RegionNorthWest + 0.2610RegionSouthEast - 0.0170RegionSouthWest + 0.0051Smoker + 0.0909BMI.Smoker - 0.0020BMI.RegionNorthWest - 0.0130BMI.RegionSouthEast - 0.0047BMI.RegionSouthWest - 0.0027Age.Children1 - 0.013Age.Children2 - 0.0087Age.Children3 - 0.0165Age.Children4 - 0.0117Age.Children5 + 0.0001Age.Bmi.SmokerNo - 0.0009Age.BMI.SmokerYes$


## Splitting the data (80:20)

```{r}
set.seed(101)
ind <- sample(2, nrow(med), replace = TRUE, prob = c(0.8, 0.2))
train <- med[ind==1,]
test <- med[ind==2,]

nrow(train)
table(train$smoker)

nrow(med)
table(med$smoker)
```

## Cross validation and Undersampling (future work)

```{r}
library(tidyverse)
library(caret) # for cross-validation methods
# Make predictions and compute the R2, RMSE and MAE

model_train <- lm(log(charges) ~ age + bmi + children + sex + region + bmi:smoker + bmi:region + age:children + age:smoker:bmi, data = train)
summary(model_train)

  
  


# RMSE for test data from model trained on training data

predictions <- predict(lm_4, test)

predictions <- exp(predictions)

data.frame( R2 = R2(predictions, test$charges),

            RMSE = RMSE(predictions, test$charges),

            MAE = MAE(predictions, test$charges))



#Undersampling on training data

under <- ovun.sample(smoker~., data=train, method = "under", N = 600)$data
table(under$smoker)

table(med$smoker)

barplot(table(under$smoker))


# Refitting the model on undersampled data

model_train_us <- lm(log(charges) ~ age + bmi + children + sex + region + bmi:smoker + bmi:region + age:children + age:smoker:bmi, data = under)
summary(model_train_us)

plot(model_train_us, 1:2)



# RMSE for test data from model trained on undersampled data
predictions <- predict(model_train_us, test)

predictions <- exp(predictions)

data.frame( R2 = R2(predictions, test$charges),

            RMSE = RMSE(predictions, test$charges),

            MAE = MAE(predictions, test$charges))
 
```


## outliers and further analysis

```{r}
# Outliers and high leverage

p <- 8
n <- nrow(med)

plot(hatvalues(lm_4), rstandard(lm_4),
xlab='Leverage', ylab='Standardized Residuals')
abline(v = 2*(p+1)/n, lty=2, col = "red") # threshold for high leverage
abline(h=c(-4,4), lty=2) # threshold for outliers


#counting the bad leverage points
med[which(abs(hatvalues(lm_4)) > (2*(p+1)/n) & abs(rstandard(lm_4)) > 4), ]
```




